---
title: "Canada's Labour Market: Past and Present"
author: "Sherry Luo and Tony Wang"
date: "03/12/2019"
output: 
  ioslides_presentation: 
    widescreen: false
    smaller: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
theme_classic()
theme_update(panel.spacing = grid::unit(0, "lines"))
library(scales)
library(viridis)
library(broom)
library(gganimate)
library(RColorBrewer)
library(plotly)
library(forcats)
library(cansim)

#Geospatial data related packages
library(ggmap)
library(sf)
library(rgdal)
library(stringr)

#Animated and sunburst related packages 
library(av)
library(sunburstR)
library(d3r)

```
## Introduction

* Topic: labour data from Statistics Canada
* Two Part exploratory analysis: 
  + Job tenure over time
  + The state of the gender pay gap today

## Job Tenure of Canada
- Based on data from Statistics Canada, we wish to explore how the length of employment changed over the period of 1976 to 2018, by age and sex. 
- Job tenure is defined as the length of time a work has been continuously employed by the same employer and is not affected if an employee changes position within the same company

## An animated pyramid plot 
- The animated pyramid plot illustrates the evolution of tenure according to gender from 1976 to 2018. 
- Selected high contrast colors and added directed labels 
- Output as MP4 to have more control of each frame and slowed down the animation with fps
- For visual purposes, assumed bin width to be equal for each bracket 

## An animated pyramid plot
<video width="750" height="450" controls>
  <source src="output_pyramid.mp4" type="video/mp4">
    </video>

## An animated pyramid plot 
* It appears that job tenure has increased for both genders and high proportion of men in each of the three categories compared to women. 
* Impact of economic trends on job tenure
  + Generally, when the economy is stable, the number of short-term jobs increase, and transition cycles follow accordingly
  + When a financial crisis strikes such as the recession in 2008 - 2012, the number of short-term jobs decrease, but the transition cycles in both the intermediate and longer-term brackets remain relatively unchanged. 
  + i.e. layoffs affect workers with lower seniority more compared to workers with higher seniority 

## Average Tenure by Sex
  *	Created interactive plot with plotly 
  *	Selected high contrast colors and to be consistent applied the same colors used in the previous plot 
  *	Interactive component (hover for more detail and able to click and hide legend)

## Average job tenure in Canada by Sex 

```{r plotlyLine_setup, message=FALSE, warning=FALSE, include=FALSE}

dta <- get_cansim(1410005101) %>% normalize_cansim_values()

#CleanData 
dta1 <- (dta
         %>% select(REF_DATE,GEO,`Job tenure`,
                    `Type of work`,Sex,`Age group`,VALUE)
         %>% filter(GEO=="Canada",
                    str_starts(`Job tenure`,"Average"),
                    str_starts(`Type of work`,"Full"),
                    str_detect(`Age group`, "to|65"),
                    str_detect(`Age group`,"44",negate = TRUE))
         %>% mutate(`Job tenure`=factor(`Job tenure`),Sex=factor(Sex),
                    REF_DATE=as.numeric(REF_DATE),`Age group`=factor(`Age group`),
                    VALUE=as.numeric(VALUE))
         %>% drop_na())

dta2 <- data.frame(dta1 
         %>% group_by(REF_DATE,Sex)
         %>% summarise(mean_tenure = mean(VALUE))
         %>% mutate(Sex = fct_relevel(Sex, c("Males","Both sexes","Females")))
         %>% rename(Year=REF_DATE,AvgTenure=mean_tenure))

# 
# (dta2 %>% plot_ly(x = ~REF_DATE,
#                  y = ~mean_tenure,
#                  split = ~Sex,
#                  #frame = ~REF_DATE,
#                  type = "scatter",
#                  mode = "lines+markers"))

plot <- print(dta2 %>% ggplot(aes(x=Year,y=AvgTenure,color=Sex))
              +geom_point()
              +geom_line()
              + labs(y="Average job tenure",x="Years",title = "Average job tenure in Canada by sex")
              + theme_minimal()
              + theme(title = element_text(size=10,face="bold"),
                      axis.title.x = element_text(size = 10,
                                                  face = "bold",vjust=0.2),
                      axis.title.y = element_text(size=10,
                                                  face="bold"),
                      axis.text.x = element_text(face="bold"),
                      axis.text.y = element_text(face="bold"))
              + scale_color_manual(values = c("#377eb8","#f89441","#cc4678")))

```

```{r averagetenurebySex, echo=FALSE}
ggplotly(plot) 

```

## Average Tenure by Sex 
  *	Examined mean tenure by gender and a more detailed trend of job tenure is exhibited  
  *	Huge gap between men and women but over time this gap narrowed as more women entered the work force around the early 90s 
  *	Mean tenure for women has been increasing steadily.
  *	After 1996, mean tenure for men has decreased compared to previous years 
  *	See impact of the recession here as well 
  *	This graph did not take account the age groups, so this plot can be misleading
  *	View age separately for each gender 
  
  
## Average Tenure by Sex and Age
  *	Selected colors with high contrast, aesthetic consistency with other plots, and to highlight the ordered age categories while trying to keep the image accessible to the visually impaired.
  *	Facets were used to allow a side-by-side comparison of trends for each gender and age group
  *	We used a linear axis as the discrepancies between groups were fairly substantial and a log scale would have been rather deceptive when it came to showing the differences between groups. 
  *	Interactive elements were used as they allow the data to be explored in greater detail without adding excessive clutter to the plot.
  *	For females, we see that for all age groups except for the 15 to 24 year age brackets, there is a slight increase in the average job tenure over time.
  *	For males aged 65 years or more, job tenure increases until 1996 and substantially decreases afterwards. We also see a similar trend in 55 to 64 year old males, though the decrease is very marginal compared to 65 years and older. 

## Sunburst Plot
  	Cumulatively from 1976 to 2018, the workforce is 40% female and 60% male. 
  -	The largest tenure bracket for both genders appears to be 1 to 5 years, though the 5 to 10 year, 10 to 20 year, and 20 plus year brackets are all fairly large as well. 
  -	We can explore the workforce composition for each sex at different job tenure levels during different years. For example, among female workers who have been at their jobs for 1 to 5 years in 2009, the majority are between 25 and 54 years old. 
  -	We can also look at the same scenario for men. In 2009, male workers who have been at their job for 1 to 5 years are overwhelmingly between 25 and 54 years old, mirroring the trend that we saw in female workers.
  
## Overall Conclusion
*	Job tenures are increasing for females, whereas job tenure for males is decreasing.  Several possible reasons for these trends are: 
    +	better workplace accommodations for women such as maternity leave, as well as possible social shifts on the role of men in parenting. 
  +	The unstable yet highly lucrative nature of work in the technology sector, which is heavily male-dominated, may partially explain the trend towards shorter job tenures for males, who may hop from one startup company to another for short periods of time before making enough money to exit the traditional full-time workforce. 
  +	Globalization may result in highly experienced workers having greater access to lucrative job positions in other countries, which would result in an outflux of workers in older age brackets.
  +	People from the silent generation may have come to retirement age in large numbers from 1997 onwards
  Choropleth
  
## Average job tenure by Age and Gender 

```{r plotlyLine1_setup, message=FALSE, warning=FALSE, include=FALSE}

###Problem with this is that age groups are mixed in...seperate this

dta3 <- data.frame(dta1 
         %>% group_by(REF_DATE,Sex,`Age group`)
         %>% summarise(mean_tenure = mean(VALUE))
         %>% rename(Year=REF_DATE,AvgTenure=mean_tenure))

# #not good
# (dta3 %>% plot_ly(x = ~REF_DATE,
#                   y = ~mean_tenure,
#                   split = ~Age.group,
#                   #frame = ~REF_DATE,
#                   type = "scatter",
#                   mode = "lines+markers"))

#display age and facet gender 
dta4 <- (dta3 %>% 
           filter(Sex !="Both sexes"))

p <- print(dta4 %>% 
             ggplot(aes(x=Year,y=AvgTenure,color=Age.group))
           + geom_point()
           + geom_line()
           + labs(y="Average job tenure",x="Years")
           + theme_minimal()
           + theme(
                   axis.title.x = element_text(size = 10,
                                               face = "bold",vjust=0.2),
                   axis.title.y = element_text(size=10,
                                               face="bold"),
                   axis.text.x = element_text(face="bold"),
                   axis.text.y = element_text(face="bold"))
           + scale_color_manual(values = c("#f89441","#cc4678","#7e03a8","#0d0887")))


p1 <- print(p 
            + facet_grid(.~Sex)
            + theme(strip.text.x = element_text(size=10,face="bold"),
                    strip.background = element_rect(fill="white")))

```

```{r averageTenure_Age_Gender, echo=FALSE}

ggplotly(p1)

```

## A Sunburst Plot
```{r sunburst_setup, message=FALSE, warning=FALSE, include=FALSE}
dta <- get_cansim(1410005101) %>% normalize_cansim_values()

#CleanData 
dta1 <- (dta
         %>% select(REF_DATE,GEO,`Job tenure`,
                    `Type of work`,Sex,`Age group`,VALUE)
         %>% filter(GEO=="Canada",str_starts(`Job tenure`,"Total",negate = TRUE),
                    str_starts(`Job tenure`,"Average",negate = TRUE),
                    str_starts(`Type of work`,"Full"),
                    Sex != "Both sexes",
                    str_detect(`Age group`, "to|65"),
                    str_detect(`Age group`,"44",negate = TRUE))
         %>% mutate(`Job tenure`=factor(`Job tenure`),Sex=factor(Sex),
                    REF_DATE=as.integer(REF_DATE),`Age group`=factor(`Age group`),
                    VALUE=as.numeric(VALUE))
         %>% drop_na())


#Reorder the job tenure
dta2 <- (mutate(dta1, `Job tenure` = fct_relevel(dta1$`Job tenure`,
                                                 c("1 to 3 months",
                                                   "4 to 6 months",
                                                   "7 to 12 months",
                                                   "13 to 60 months",
                                                   "61 to 120 months",   
                                                   "121 to 240 months",
                                                  "241 months or more"))))

#Recoding labels for clarity
dta3 <- mutate(dta2, `Job tenure` = (dta2$`Job tenure`  
                                     %>% fct_recode(`Job Tenure: 1 to 3 months`="1 to 3 months",
                                                    `Job Tenure: 4 to 6 months` ="4 to 6 months",
                                                    `Job Tenure: 7 to 12 months`="7 to 12 months",
                                                    `Job Tenure: 1 to 5 years`="13 to 60 months",
                                                    `Job Tenure: 5 to 10 years`="61 to 120 months",
                                                    `Job Tenure: 10 to 20 years`="121 to 240 months",
                                                    `Job Tenure: 20 years plus`= "241 months or more")))

dta31 <- mutate(dta3, `Age group`= (dta3$`Age group` 
                            %>% fct_recode(`Age Group: 15 to 24 years`="15 to 24 years",
                            `Age Group: 25 to 54 years` = "25 to 54 years",
                            `Age Group: 55 to 64 years`="55 to 64 years",
                            `Age Group: 65 years and over`="65 years and over")))


dta4 <- data.frame((dta31 
                    %>% mutate(REF_DATE=factor(REF_DATE))
                    %>% select(`Sex`,`Job tenure`,REF_DATE,`Age group`,VALUE)
                    %>% unite(seq,`Sex`:`Age group`,sep="-")
                    %>% select(seq,VALUE)))

```

http://rpubs.com/sluoo/sunburst

```{r sunburstPlot}
sunplot1 <- sund2b(dta4,
                  rootLabel = "Sex",
                  colors = htmlwidgets::JS("d3.scaleOrdinal(d3.schemeCategory20b)"),
                  tooltip =  sund2bTooltip(followMouse = TRUE,
                                           html = htmlwidgets::JS("function(nodedata, size, percent) {
  return '<span style=\"font-weight: bold;\">' + nodedata.name + '</span>' + ' ' + size
}")) 
)

sunplot1

```



## Pay Gaps, Industry, and Geography
* When talking about job data, one thing that frequently comes up is the gender wage gap
* Common saying is that women get paid 70 cents for every dollar men get paid
* Problem is that pay depends on many different factors
* Among people with the same educational level and the same field of study, is there a major gender pay gap for different parts of Canada?

## Pay Gaps, Industry, and Geography
* We combined two subsets of the 2016 Census Data
  + Employment Income by Major Field of Study (98-400-X2016280)
  + Field of Study Groupings for Age Groups, Sex, and educational attainment (98-402-X2016010-T4)
* We then calculated the proportion of females in the workforce using the count data of each cohort
* Cleaned data was used to create a faceted map
* ggplotly was used to make the facet plot interactive
* Tooltips of detailed information were added using "dummy" aesthetics in ggplot

## Pay Gaps, Industry, and Geography

```{r geospatial_setup, message=FALSE, warning=FALSE, include=FALSE}

##Wage Data File by Industry
##Custom extraction from original 6GB census data
wages <- read_csv("Utility/wageData.csv")

## Unzip and read CSV for Industry by Gender (Note: Parsing failures are blank rows in the data)
## Many, many columns to rename...
## Drop useless columns and NA rows
source <- read_csv("Utility/98-402-X2016010-T4-CANPR-eng.csv") 
source_renamed <- (source %>% rename("GeoName"="Geographic name", 
                                     "AreaTotal"="Total – STEM and BHASE (non-STEM) groupings - Classification of Instructional Programs (CIP) 2016 (2016 counts)",
                                     "Science and science technology"="STEM fields of study - Science and science technology (2016 counts)",
                                     "Engineering and engineering technology"="STEM fields of study - Engineering and engineering technology (2016 counts)",
                                     "Mathematics and computer and information science"="STEM fields of study - Mathematics and computer and information science (2016 counts)",
                                     "Business and administration"="BHASE (non-STEM) fields of study - Business and administration (2016 counts)",
                                     "Arts and humanities"="BHASE (non-STEM) fields of study - Arts and humanities (2016 counts)",
                                     "Social and behavioural sciences"="BHASE (non-STEM) fields of study - Social and behavioural sciences (2016 counts)",
                                     "Legal professions and studies"="BHASE (non-STEM) fields of study - Legal professions and studies (2016 counts)",
                                     "Health care"="BHASE (non-STEM) fields of study - Health care (2016 counts)",
                                     "Education and teaching"="BHASE (non-STEM) fields of study - Education and teaching (2016 counts)",
                                     "Trades, services, natural resources and conservation"="BHASE (non-STEM) fields of study - Trades, services, natural resources and conservation (2016 counts)",
                                     "TotalPercent"="Total – STEM and BHASE (non-STEM) groupings - Classification of Instructional Programs (CIP) 2016 (% distribution 2016)"
)
%>% drop_na()
%>% filter(GeoName!="Canada")
%>% filter(Age=="All ages, 15-plus")
%>% dplyr::select(-c("Geographic code", "Age", "Global non-response rate", "Data quality flag", 19:29))
)

## Join wage data with demographic data
merged <- (source_renamed %>% gather(c(5:14), key="Industry", value="Cohort Size") 
%>% left_join(wages))

##Numbers for both genders
bothsexes <- (merged
              %>% filter(Sex== "Both sexes")
              %>% rename("Median Income (Both Genders)"="Median Income")
              %>% dplyr::select(-c("Sex"))
)

##Numbers for females only
females <- (merged
            %>% filter(Sex== "Female")
            %>% rename("Median Income (Female)"="Median Income",
                       "FemaleGenderCount"="Cohort Size")
            %>% dplyr::select(-c("Sex", "AreaTotal"))
)

##Numbers for males only
males <- (merged
            %>% filter(Sex== "Male")
            %>% rename("Median Income (Male)"="Median Income")
          %>% dplyr::select(-c("Cohort Size", "Sex", "AreaTotal"))
)


##Recombine Tables and Calculate Female Proportions
##Filter out Most Frequent Industries per Jurastiction
femaleprop <- (left_join(bothsexes,females) %>% left_join(males)
               %>% mutate(FemalePercent = round(100*(FemaleGenderCount/`Cohort Size`), digits=0))
)

##Extract most common industry by education level, based on femaleprop table
##Create Five Tables For Different Education Levels
##Within each table, extract top value for each province, then rejoin all at the end
##Not ideal but this level of striation requires some drastic measures to clean

College <- (femaleprop %>% filter(Education=="College, CEGEP or other non-university certificate or diploma") 
            %>% group_by(GeoName) %>% slice(which.max(`Cohort Size`)))
UniCert <- (femaleprop %>% filter(Education=="University certificate, diploma or degree at bachelor level or above")
            %>% group_by(GeoName) %>% slice(which.max(`Cohort Size`)))
Bach <- (femaleprop %>% filter(Education=="Bachelor's degree")
         %>% group_by(GeoName) %>% slice(which.max(`Cohort Size`)))
Master <- (femaleprop %>% filter(Education=="Master's degree")
           %>% group_by(GeoName) %>% slice(which.max(`Cohort Size`)))
PhD <- (femaleprop %>% filter(Education=="Earned doctorate")
        %>% group_by(GeoName) %>% slice(which.max(`Cohort Size`)))

#Final Cleaned Data Table
dtafin <- (bind_rows(College, UniCert, Bach, Master, PhD, id=NULL))
##Factor Education Column (essentially for facet reordering)
dtafin$Education <- factor(dtafin$Education, levels = c("College, CEGEP or other non-university certificate or diploma",
                                                        "University certificate, diploma or degree at bachelor level or above",
                                                        "Bachelor's degree",
                                                        "Master's degree",
                                                        "Earned doctorate"), ordered = TRUE)             
## If no Income data is available, replace value with "No Data"
dtafin$`Median Income (Female)`[which(dtafin$`Median Income (Female)`==0)] <- "No Data"
dtafin$`Median Income (Male)`[which(dtafin$`Median Income (Male)`==0)] <- "No Data"

####################################################

## Load Shape File, Join Shape File with Data 
## Use digital boundary file for performance
canada_shape <- st_read("Shape Files/Digital Boundary/lpr_000a16a_e.shp")
joined_dta <- left_join(canada_shape, dtafin, by=c("PRENAME"="GeoName")) %>% dplyr::rename("Jurisdiction"="PRENAME")

##Change strip text names without renaming data
Facet_names <- c(
  `College, CEGEP or other non-university certificate or diploma` = "College and Other",
  `University certificate, diploma or degree at bachelor level or above` = "University Certificate",
  `Bachelor's degree` = "Bachelor's degree",
  `Master's degree` = "Master's degree",
  `Earned doctorate` = "Earned doctorate"
)

#Create ggplot with dummy aesthetics for later use
plot <- (ggplot(joined_dta)
         + geom_sf(aes(common=Jurisdiction, 
                       fill=Industry, label1=`Cohort Size`, 
                       label2=FemalePercent, 
                       label3=`Median Income (Female)`,
                       label4=`Median Income (Male)`,
                       geometry=geometry))
         + scale_fill_viridis(discrete=TRUE, option="plasma")
         + xlab("Longitude")
         + ylab("Latitude")
         + facet_wrap(~Education, labeller = as_labeller(Facet_names))
         + theme_classic()
         + guides(alpha=FALSE)
         + theme(legend.position=c(0.83, 0.25), 
                 panel.spacing = unit(0.5, "lines"),
                 axis.text = element_blank(),
                 axis.ticks=element_blank(),
                 legend.title = element_blank(),
                 axis.title.x = element_blank(),
                 axis.title.y = element_blank(),
                 )
)

#Create ggplot with dummy aesthetics for later use
plot <- (ggplot(joined_dta)
         + geom_sf(aes(common=Jurisdiction, 
                       fill=Industry, label1=`Cohort Size`, 
                       label2=FemalePercent, 
                       label3=`Median Income (Female)`,
                       label4=`Median Income (Male)`,
                       geometry=geometry))
         + scale_fill_viridis(discrete=TRUE, option="plasma")
         + xlab("Longitude")
         + ylab("Latitude")
         + facet_wrap(~Education, labeller = as_labeller(Facet_names))
         + theme_classic()
         + guides(alpha=FALSE)
         + theme(legend.position=c(0.83, 0.25), 
                 panel.spacing = unit(0.5, "lines"),
                 axis.text = element_blank(),
                 axis.ticks=element_blank(),
                 legend.title = element_blank(),
                 axis.title.x = element_blank(),
                 axis.title.y = element_blank(),
                 )
)
```

```{r geospatial, message=FALSE, warning=FALSE}
(ggplotly(plot, tooltip=c("common", "label1", "label2", "label3", "label4"), width=800, height=450) 
  %>% add_annotations(text="Predominant Field for Workforce Cohort",
                      xref="paper", yref="paper",
                      x=0.7, xanchor="left",
                      y=0.47, yanchor="top",
                      legendtitle=TRUE, showarrow=FALSE)
  %>% layout(autosize=F, legend=list(x=0.7, y=0.025, yanchor="bottom"), hovermode = "closest") 
)
```

## Overall Conclusions
* Canada's workforce:
  + Has changed a lot over time demographically
  +
  + Exhibits a gender pay gap