---
title: "Canada's Labour Market: Past and Present"
author: "Sherry Luo and Tony Wang"
date: "03/12/2019"
output: 
  ioslides_presentation: 
    widescreen: false
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(shiny)
```
## Introduction

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Age Distribution

- Pyramid Plot
- Bullet 2
- Bullet 3

## Job tenure by Year, Age and Gender
- Sunburst and Animated Dot Plots


## Pay Gaps, Industry, and Geography

- Faceted Choropleth
- Bullet 2
- Bullet 3

## Pay Gaps, Industry, and Geography

```{r geospatial, message=FALSE, warning=FALSE}
includeHTML("Presentation/choropleth.html") 
```

